Supplemental Material for ``Assessing United States county-level exposure for research on tropical cyclones and human health"

### 2019-11-07 writing/darlings_intro.tex
Make more edits

>>>>

, either to assess average
exposure patterns for tropical cyclones (e.g., \cite{zandbergen2009}) or

These analyses can also help
researchers compare and interpret results from previous epidemiological studies
that have assessed tropical cyclone exposure in different ways, helping to
[synthesize] previous epidemiologic results to form a better picture of how
these storms can affect human health.

Tropical cyclones vary
dramatically in size: \ac{US} tropical cyclones have been observed with radii
to maximum winds as small as~20~\si{\kilo\metre} and as large
as~200~\si{\kilo\metre}~\parencite{mallin2006, quiring2011variations}.

### 2019-11-07 writing/darlings_results.tex
Make more edits

>>>>

For most tropical cyclones,
distance-based county-level exposure assessment was, at best, in moderate
agreement with exposure assessments based on the four hazard-based metrics.

\subsection*{Case study}

We conducted a case study comparing estimates of physical exposure to tropical
cyclones based on each exposure metric.  Figure~\ref{fig:topelecdependexposure}
shows the study counties with the highest expected rates of physical exposure
to tropical cyclones among electricity-dependent Medicare beneficiaries based
on each metric. A few counties (Miami-Dade County, FL; Harris County, TX) were
ranked in the top three in expected yearly rate of physical exposure under
almost all exposure metrics (Figure~\ref{fig:topelecdependexposure}). However,
there were key differences across exposure metrics. For example, highly-ranked
based on wind- and tornado-based metrics were in Florida, while over half of
the highly-ranked counties based on the flood-based metric were in Mid-Atlantic
states (PA, NY).

### 2019-11-07 writing/ehp_discussion.tex
Make more edits

Epidemiological studies can help characterize which health risks are elevated
during disasters, to what degree, and to whom [Ibrahim editorial; Noji paper].
\parencite{ibrahim2005unfortunate, noji2005disasters}.  As a result, these studies help improve
disaster preparedness and response \parencite{noji2005disasters}.  However,
tropical cyclones are multi-hazard events, making it complicated to measure
exposure.  Here, we provide open-source county-level data on multiple tropical
cyclone exposure metrics and explore limitations in that data.

Further, we find that county-level tropical cyclone exposure assessments vary
substantially when using different metrics, based either on storm hazards or on
how close the storm's central track came to the community. Our results can
inform exposure assessment for future county-level studies of the health risks
and impacts associated with tropical cyclones exposures, and the open-source
data and software we have created provides multi-hazard, county-level exposure
data for such studies.

@@

We investigated how well the open-source data correspond with
data from other available sources
(Figures~\ref{fig:raincomparison}\,--\,\ref{fig:floodcomparison} and~S2). While
generally in agreement, there are a few caveats that should be considered when
using these data. The rainfall data are generally
well-correlated with ground-based observations, but may sometimes
oversmooth very high rainfall values compared to ground-based observations
(Figure~\ref{fig:raincomparison}). When rainfall data is used to create binary
exposure classifications, this disagreement is unlikely to influence results,
as both data sources agree in identifying these as storms with high rainfall,
but would be important to consider for cases that include rainfall as a
continuous measurement.

@@

For the flooding data, we found that flood event status determined based
on the NOAA Storm Events listings often agreed with measurements from \ac{USGS}
streamgages, with a flood event more likely to be listed if a storm elevated
streamflow across the county (Figure~\ref{fig:floodcomparison}).  However, there
are differences between the two flooding datasets, and these highlight both the
difficulty of measuring flood exposure at the county level and inherent
challenges in using data from a storm event database for epidemiologic exposure
assessment.

For example, there was one storm in Fulton County, GA, for which there was high
streamflow but not an associated flood event listing (Hurricane Ida, 2009).
This storm occurred in November 2009, following a month with historic rainfall
and flooding in Georgia \parencite{shepherd2011overview}.  Therefore, in this
case the flooding associated with Ida was incorporated into an ongoing flood
event listing, with a start date well before the five-day window we used to
temporally match storm event listings with tropical cyclone tracks.

@@

There are further limitations for the flood and tornado data---these data came
from the \ac{NOAA} Storm Events database, which, while a widely-used database
of events maintained by \ac{NOAA}, is based on reports, and so may be prone to
underreporting~\parencite{Ashley2008flood, Curran2000}, especially in
sparsely-populated areas~\parencite{Witt1998, Ashley2007} [Simmons and Sutter
ref], as well as to other reporting errors.

While these are important caveats for the data, we selected these data sources
as among the best currently available for measuring each of these hazards
consistently and comprehensively at a multi-county, multi-year scale. In
addition to providing tropical cyclone exposure metrics for individual hazards,
this dataset and its associated software allow users not only to access
measurements for single hazards, but also to create tropical cyclone exposure
profiles based on multiple hazards or craft exposure indices that combine
hazard metrics~\parencite{chakraborty2005population, peduzzi2009assessing}.
This ability can be critical, as different hazards of tropical cyclones often
act synergistically in causing impacts~\parencite{smith2009}.

@@

We found differences in geographic patterns in average exposures to different
tropical cyclone hazards (Figure~\ref{fig:averageexposure}). These patterns
were not unexpected based on what is known about tropical cyclone hazards, but
still highlight variations that are critical to consider in designing studies
and statistical analysis for tropical cyclone epidemiology. Further, they
demonstrate the need for multi-hazard exposure datasets for tropical cyclone
epidemiology.

Tropical cyclone wind exposures had a strong coastal pattern, which is
consistent with the dramatic decrease in wind intensity that typically
characterizes the landfall of tropical cyclones.  Tropical cyclone rain
exposures tended to extended further inland compared to wind exposures, up to
the Appalachian mountains. This agrees with previous research indicating that
the Appalachian mountains' topography both enhances precipitation during
tropical cyclones and provides hydrological conditions for severe
flooding~\parencite{rees2001}.  Almost all tropical cyclone tornado exposures
were in southern coastal states, consistent with previous evidence that most noteworthy
tropical cyclone-related tornadoes occur on the right-side of the tropical
cyclone track in Atlantic-basin \ac{US} storms~\parencite{moore2012}.  The
exposure averages we calculated are somewhat limited as estimates of long-term
frequencies, as tropical cyclones follow decadal
patterns~\parencite{kossin2007more} likely not adequately captured in the
available data, which covered less than 20 years. However, exposure assessments
spanning over a century have found similar patterns for distance- and
wind-based metrics~\textcite{zandbergen2009, kruk2010} [double-check that refs
found similar patterns].

@@

We found that tropical cyclones tended to bring different hazards to 
different counties and also that agreement was typically low between
distance-based tropical cyclone exposure assessment and each of the single
hazard-specific exposures (Figures~\ref{fig:ivanexposure}\,--\,
\ref{fig:jaccard} and S4\,--\,S5). This suggests that the use of a
distance-based metric to assess exposure to any of these hazards, or the use of
measurements from one hazard as a proxy for exposure to any of the other
hazards considered, would often be problematic, introducing exposure
misclassification.

These findings align with previous results from atmospheric science and related
fields on the characteristics of tropical cyclones. While tropical cyclone
rainfall and windspeed can be well-correlated when the tropical cyclone is
over water~\parencite{cerveny2000}, this relationship often weakens once the
hurricane has made landfall~\parencite{jiang2008}.  Fast-moving tropical
cyclones bring higher risks of dangerous winds inland~\parencite{kruk2010},
while slow-moving tropical cyclones are likely to bring more
rain~\parencite{rappaport2000} and cause more damage because of sustained
hazardous conditions~\parencite{rezapour2014}. Further, while the likelihood
and extent of flooding during a tropical cyclone is related to the tropical
cyclone's rainfall, it is also driven by factors like top soil saturation and
the structure of the water basin's drainage network~\parencite{chen2015,
rees2001}.

Therefore for most tropical cyclones, since exposure assessments often differ
across storm hazards, as well as between each storm hazard and a distance-based
proxy measurement, exposure misclassification is a potential risk. For some
studies, such exposure misclassification might plausibly be differential (i.e.,
associated with the outcome of interest or with factors associated with risk of
the outcome of interest).  For example, tropical cyclone wind exposures tend to
be concentrated in counties near the coast, while tropical cyclone wind
exposures sometimes extend well inland.  If the etiologically-relevant exposure
for a health outcome is extreme rainfall but exposure is classified based on
measurements of wind, the probability of being misclassified as unexposed would
be higher in inland counties, while the probability of being misclassified as
exposed would be higher in coastal counties. If coastal counties differ from
inland counties in either the outcome of interest or in factors associated with
risk of that outcome, exposure misclassification would be
differential~\parencite{savitz2016interpreting}.  Such differential exposure
misclassification could bias estimates of tropical cyclone effects either
towards the null (estimating a lower or null association compared to the true
association that exists) or away from the null (estimating a larger association
than actually exists)~\parencite{savitz2016interpreting, armstrong1998effect}.
[Variation across studies in how exposure is assessed?]

However, we did find a small set of tropical cyclones for which for which
agreement was high across several single-hazard exposure assessments (e.g.,
Floyd in~1999, Irene in 2011, Hannah in~2008, Bertha in~1996; Ernesto in~2006
Figure~\ref{fig:jaccard}).  Hurricanes Floyd in~1999 (Figure~S4) and Irene
in~2011 both made their first \ac{US} landfall in North Carolina at minor
hurricane windspeeds (Category~2 and~1, respectively) and then skimmed the
eastern coast of the \ac{US} north through New England, bringing substantial
rainfall to much of the eastern coast from North Carolina north and causing
extensive inland flooding in North Carolina (Floyd) and New England
(Irene)~\parencite{avila2013atlantic, lawrence2000atlantic}.  Hurricanes Hannah
in~2008, Bertha in~1996, and Ernesto in~2006 also followed the eastern
coastline. For this set of storms, the tropical cyclones' persistent proximity
to water may have helped maintain wind speeds in similar patterns to rain and
distance-based exposure assessments, resulting in more similarities across
exposure assessments compared to the other tropical cyclones we considered.
For these storms, it may be possible to assess exposure to multiple hazards of
the storm using a single metric, in some cases even a proxy like the distance
between a county and the storm's track.  However, for these storms it may be
difficult to try to untangle the contribution of each hazard to the overall
effect of the storm, given that several hazards have similar geographical
patterns.

\subsection*{Case study}

The case study illustrates how differences in exposure assessment across storm
exposure metrics can result in differing conclusions. In this case, we found
differences in estimates of expected rates of physical exposure to tropical
cyclones among a population when exposure assessment was based on different
storm exposure metrics (Figure~\ref{fig:topelecdependexposure}).  These results
are informative because previous assessments of physical exposure to tropical
cyclones have often concentrated on storm winds and storm proximity. For
example, a global study of exposure and vulnerability to natural hazards used a
combination of windspeed above a certain threshold and distance from the
storm's track to assess exposure~\parencite{peduzzi2009assessing}.  While many
storm impacts might be most strongly linked to wind hazards, there is a growing
recognition of the potential risks of adverse health outcomes and property
damage from rain- and flood-related tropical cyclone hazards in the
\ac{US}~\parencite{smith2009}.  Assessments based solely on wind and distance
from the storm track therefore may underemphasize potentially dangerous
exposures to certain regions of the \ac{US}, particularly inland locations.  In
fact, based on our assessment of physical exposures to tropical cyclones among
electricity-dependent Medicare beneficiaries, the highest expected physical
exposure under any metric was measured for Philadelphia County, PA, under the
flood metric (Figure~\ref{fig:topelecdependexposure}).  While this county has a
similar-sized population of electricity-dependent Medicare beneficiaries as
Miami-Dade County, FL (Figure~\ref{fig:topelecdependexposure}), the inland
county of Philadelphia has frequent storm-related flood events, which results
in a very high expected physical exposure among electricity-dependent Medicare
beneficiaries in the county. This exposure is poorly captured by most of the
other exposure metrics, under which Philadelphia County is not included in the
top ten counties for expected rates of physical
exposure~(Figure~\ref{fig:topelecdependexposure}).

@@

Previous research has highlighted the range of impacts that tropical cyclones
can have in \ac{US} communities. However, studies of tropical cyclone exposures
and their impacts have varied widely in how they assess exposure, and there was
not previously a widely available public dataset that captured county-level
exposure to multiple storm hazards. Here we provide open-source tropical
cyclone hazard-related measurements for a large collection of historical
Atlantic-basin cyclones, aggregated at the county level to align with the
spatial scale at which much health-related data is collected.

Further, we investigated agreement in exposure assessment across these metrics
when applied individually. We found large differences in which counties are
exposed to different hazards of tropical cyclones and that distance is, at
best, a moderate, and often a very poor, surrogate for exposure to the specific
tropical cyclone hazards of high wind, extreme rainfall, flooding, and
tornadoes. Use of distance as a surrogate for any of these hazards, or exposure
assessment based on one hazard when the pathway for health impacts is in part
or full through another storm hazard, could lead to exposure misclassification.
In the case of tropical cyclone risk and impact studies, including
epidemiological studies, this would result in biased estimates.  Our findings
highlight the importance of clarifying the potential pathway from tropical
cyclone hazards to health impacts when conducting tropical cyclone
epidemiological studies, and then basing exposure assessment on these hazards.

>>>>

Epidemiological studies can help characterize which health risks are elevated
during disasters, to what degree, and to
whom~\parencite{ibrahim2005unfortunate, noji2005disasters}.  As a result, these
studies help improve disaster preparedness and
response~\parencite{noji2005disasters}.  However, tropical cyclones are
multi-hazard events, making it complicated to measure exposure and so to
conduct multi-community, multi-year studies leveraging large administrative
datasets.  Here, we provide open-source county-level data on multiple tropical
cyclone exposure metrics and explore limitations in that data.

Further, we explore patterns in storm exposure classifications based on
different metrics, and we find that county-level tropical cyclone exposure
assessments vary substantially when using different metrics, based either on
storm hazards or on how close the storm's central track came to the community.
Our results can inform exposure assessment for future county-level studies of
the health risks and impacts associated with tropical cyclones exposures as
well as insights to help in designing studies.  The open-source data and
software we have created provides multi-hazard, county-level exposure data for
such studies.

@@

We investigated how well the open-source data correspond with data from other
potential sources
(Figures~\ref{fig:raincomparison}\,--\,\ref{fig:floodcomparison} and~S2). While
generally in agreement, there are some caveats. The rainfall data are generally
well-correlated with ground-based observations, but may sometimes oversmooth
very high rainfall values compared to ground-based observations
(Figure~\ref{fig:raincomparison}). When rainfall data is used to create binary
exposure classifications, this disagreement is unlikely to influence results,
as both data sources agree in identifying these as storms with high rainfall,
but would be important to consider for cases that include rainfall as a
continuous measurement.

@@

For the flooding data, we found that flood event status determined based on the
NOAA Storm Events listings typically agreed with measurements from \ac{USGS}
streamgages, with a flood event more likely to be listed if a storm elevated
streamflow across the county (Figure~\ref{fig:floodcomparison}).  However,
there are differences between the two flooding datasets, and these highlight
both the difficulty of measuring flood exposure at the county level and
inherent challenges in using data from a storm event database for epidemiologic
exposure assessment.  For example, there was one storm in Fulton County, GA,
for which there was high streamflow but not an associated flood event listing
(Hurricane Ida, 2009).  This storm occurred in November 2009, following a month
with historic rainfall and flooding in
Georgia~\parencite{shepherd2011overview}.  In this case the flooding
associated with Ida was incorporated into an ongoing flood event listing, with
a start date well before the five-day window we used to temporally match storm
event listings with tropical cyclone tracks.

@@

There are further limitations for the flood and tornado data---these data came
from the \ac{NOAA} Storm Events database, which, while a widely-used database
of events maintained by \ac{NOAA}, is based on reports, and so may be prone to
underreporting~\parencite{Ashley2008flood, Curran2000}, especially in
sparsely-populated areas~\parencite{Witt1998, Ashley2007}, as well as to other
reporting errors.

While these are important caveats for the data, we selected these data sources
as among the best currently available for measuring each of these hazards
consistently and comprehensively at a multi-county, multi-year scale. In
addition to providing tropical cyclone exposure metrics for individual hazards,
this dataset and its associated software allow users not only to access
measurements for single hazards, but also to create tropical cyclone exposure
profiles based on multiple hazards or to craft exposure indices that combine
hazard metrics~\parencite{chakraborty2005population, peduzzi2009assessing}.
This ability can be critical, as different hazards of tropical cyclones can
act synergistically in causing impacts~\parencite{smith2009}.

@@

We found average exposures to different tropical cyclone hazards differed
geographically (Figure~\ref{fig:averageexposure}). These patterns were not
unexpected based on what is known about tropical cyclone hazards, but still
highlight variations that are critical to consider in designing studies and
statistical analysis for tropical cyclone epidemiology. Further, they
demonstrate the need for multi-hazard exposure datasets for tropical cyclone
epidemiology, especially in capturing inland risks.

Tropical cyclone wind exposures had a strong coastal pattern, which is
consistent with the dramatic decrease in wind intensity that typically
characterizes the landfall of tropical cyclones.  Tropical cyclone rain
exposures tended to extended further inland compared to wind exposures, up to
the Appalachian mountains. This agrees with previous research indicating that
the Appalachian mountains' topography both enhances precipitation during
tropical cyclones and provides hydrological conditions for severe
flooding~\parencite{rees2001}.  Almost all tropical cyclone tornado exposures
were in southern coastal states, consistent with previous evidence that most
noteworthy tropical cyclone-related tornadoes occur on the right-side of the
tropical cyclone track in Atlantic-basin \ac{US} storms~\parencite{moore2012}.
The exposure averages we calculated are somewhat limited as estimates of
long-term frequencies, as tropical cyclones follow decadal
patterns~\parencite{kossin2007more} likely not adequately captured in the
available data, which covered less than 20 years. However, exposure assessments
spanning over a century have found similar patterns for distance- and
wind-based metrics~\textcite{zandbergen2009, kruk2010} \textbf{[double-check
that refs found similar patterns]}.

@@

We found that tropical cyclones tended to bring different hazards to different
counties and also that agreement was typically low between distance-based
tropical cyclone exposure assessment and each of the single hazard-specific
exposures, as well as between pairs of hazard-specific metrics
(Figures~\ref{fig:ivanexposure}\,--\, \ref{fig:jaccard} and S4\,--\,S5).  These
findings align with previous results from atmospheric science and related
fields on the characteristics of tropical cyclones. While tropical cyclone
rainfall and windspeed can be well-correlated when the tropical cyclone is over
water~\parencite{cerveny2000}, this relationship often weakens once the
hurricane has made landfall~\parencite{jiang2008}.  Fast-moving tropical
cyclones bring higher risks of dangerous winds inland~\parencite{kruk2010},
while slow-moving tropical cyclones are likely to bring more
rain~\parencite{rappaport2000} and cause more damage because of sustained
hazardous conditions~\parencite{rezapour2014}. Further, while the likelihood
and extent of flooding during a tropical cyclone is related to the tropical
cyclone's rainfall, it is also driven by factors like top soil saturation and
the structure of the water basin's drainage network~\parencite{chen2015,
rees2001}.

This suggests that the use of a distance-based metric to assess exposure to any
of these hazards, or the use of measurements from one hazard as a proxy for
exposure to any of the other hazards considered, would often be problematic,
introducing exposure misclassification. \textbf{[Compare with Grabich results?]}
For some studies, such exposure misclassification might plausibly be
differential (i.e., associated with the outcome of interest or with factors
associated with risk of the outcome of interest).  For example, tropical
cyclone wind exposures tend to be concentrated in counties near the coast,
while tropical cyclone wind exposures sometimes extend well inland.  If the
etiologically-relevant exposure for a health outcome is extreme rainfall but
exposure is classified based on measurements of wind, the probability of being
misclassified as unexposed would be higher in inland counties, while the
probability of being misclassified as exposed would be higher in coastal
counties. If coastal counties differ from inland counties in either the outcome
of interest or in factors associated with risk of that outcome, exposure
misclassification would be differential~\parencite{savitz2016interpreting}.
Such differential exposure misclassification could bias estimates of tropical
cyclone effects either towards the null (estimating a lower or null association
compared to the true association that exists) or away from the null (estimating
a larger association than actually exists)~\parencite{savitz2016interpreting,
armstrong1998effect}.

We did, however, find a small set of tropical cyclones for which for
which agreement was high across several single-hazard exposure assessments
(e.g., Floyd in~1999, Irene in 2011, Hannah in~2008, Bertha in~1996; Ernesto
in~2006 Figure~\ref{fig:jaccard}).  Hurricanes Floyd in~1999 (Figure~S4) and
Irene in~2011 both made their first \ac{US} landfall in North Carolina at minor
hurricane windspeeds (Category~2 and~1, respectively) and then skimmed the
eastern coast of the \ac{US} north through New England, bringing substantial
rainfall to much of the eastern coast from North Carolina north and causing
extensive inland flooding in North Carolina (Floyd) and New England
(Irene)~\parencite{avila2013atlantic, lawrence2000atlantic}.  Hurricanes Hannah
in~2008, Bertha in~1996, and Ernesto in~2006 also followed the eastern
coastline. For these storms, the tropical cyclones' persistent proximity
to water may have helped maintain wind speeds in similar patterns to rain and
distance-based exposure assessments, resulting in more similarities across
exposure assessments compared to the other tropical cyclones we considered.
For these storms, it may be possible to assess exposure to multiple hazards of
the storm using a single metric, in some cases even a proxy like the distance
between a county and the storm's track.  However, for these storms it may be
difficult to try to untangle the contribution of each hazard to the overall
effect of the storm, given that several hazards have similar geographical
patterns.

\subsection*{Limitations}

The dataset presented here does have several limitations, in addition to the
caveats previously discussed. First, the dataset is not comprehensive of all
possible tropical cyclone hazards. For example, coastal counties can experience
dangerous storm surge, which is not specifically covered in this dataset
(although some resulting flooding would be captured). We are exploring ways to
include this in future versions of the dataset, although to date we have
focused on exposures that could affect any county, whether inland or coastal.
Second, this data is aggregated to the county level. Many health outcome
datasets are aggregated at this level, but some may be aggregated at a finer
spatial resolution (e.g., Census tract or ZIP code). We have published the wind
model we used to create this dataset as its own open-source R package
\parencite{stormwindmodel}, and this model can be used to model
storm-associated winds at a finer spatial resolution; however, measurements of
other hazards cannot similarly be re-scaled through tools we provide. Finally,
we provide these data and associated software tools through R packages, and so
some experience in the R programming language is required to make full use of
them. However, R is currently a popular programming language for environmental
epidemiology, allowing the data to reach a large audience, and we are exploring
options to create a web application using the Shiny platform to allow more
user-friendly web-based access of the data \parencite{shiny2019}.

@@

To conduct epidemiological studies that span multiple communities and tropical 
cyclones, it is critical to have consistent and comprehensive measurements of 
community-level exposure to hazards of the storm. Here we have created and 
shared a dataset that provides these data for counties in the United States
over multiple years. Despite some limitations in this dataset, it provides 
a powerful tool for expanding tropical cyclone epidemiology studies to more
extensively leverage existing administrative and vital data, allowing researchers
to improve our understanding of how these storms affect community-wide risk of 
numerous health outcomes, including outcomes that are also common outside of storm 
periods and so may be hard to attribute on a case-by-case basis to a storm. 
Further, this dataset provides hazard measurements that are comparable across
communities and storms, which allows epidemiological researchers to design studies
to explore how health risks are modified by characteristics of both the storms and
the communities that are hit. The data are given in an open-source format, along 
with associated software tools, which allows them to be freely used and for others
to explore all associated code and to contribute additions through platforms like 
GitHub.

Based on our analysis in this paper, these data sources are typically in agreement
with other sources of data available to characterize storm-associated hazards
(e.g., ground-based monitors, streamgages, post-storm wind radii estimates). 
However, researchers who are planning to use the data should explore the analyses
presented in this paper to understand the strengths and weaknesses of the data. 
Our analyses also indicate that a storm often brings different combinations of 
hazards to different counties along or near its path. Our results in comparing 
storm hazards suggests that county-level storm exposure is not well-characterized 
by the closest distance that a storm's central track came to a county, and that
exposure to one storm hazard within a county (e.g., severe winds) does not imply 
exposure to other hazards (e.g., excessive rainfall, flooding). As a result, it is 
critical that researchers consider which storm hazards are likely on the causal 
pathway for the outcomes they are studying, and to characterize storm exposure
in a way that captures those specific hazards.
